Relational Parametricity for Algebraically Indexed Types

### 2012-06-06 paper/geometry-twocolumn.tex
More updates to the motivating examples section

\subsection{2-Dimensional Geometry and Invariance}

When writing programs that manipulate $n$-dimensional geometric data,
the basic data structure is the $n$-tuple of real numbers. In the
two-dimensional case, tuples $\vec{v} = (x,y)$ can be called upon to
represent both \emph{points} -- offsets from some origin -- and
\emph{vectors} -- offsets in their own right. The notions of point and
vector are very different despite their common representation. By
representing a point concretely as an offset from some origin, we
inadvertently fix that origin as privileged.
% However, as made clear by Gallieo, many of the laws of physics are
% invariant under change of the frame of reference: if we change where
% we think the origin is, then we do not change the fundamental laws.
Putting this in concrete terms, if we have two frames of reference,
one with $(0,0)$ as the origin, and one with $(10,20)$ as the origin,
then the tuple $(1,1)$ with respect to the first frame and $(11,21)$
with respect to the second frame represent the \emph{same} point. It
is merely an artifact of our representation of points as pairs of
numbers that these appear to be different. When we write programs that
manipulate data that represents points, we ought to follow the general
principle that our programs are invariant with respect to changes in
the choice of origin. On the other hand, vectors are not intended to
be regarded as invariant under change of choice of origin. Vectors
represent offsets, or movements, and the vector $(0,0)$ always
represents the zero offset.
% However, they are intended to be invariant under change of basis: the
% vector $(2,3)$, with respect to the basis $(1,0),(0,1)$, and the
% vector $(1,1)$, with respect to the basis $(2,0),(0,3)$, should be
% regarded as the same vector. Invariance under change of basis also
% applies to the representation of points. We therefore refine the
% principle of invariance under change of origin for points to also
% include invariant under change of basis for points and vectors.

@@

\textbf{FIXME:} talk about the standard way of separating the two
concepts using abstract types.

Invariance under change of representation immediately recalls
Reynolds' fable concerning two professors teaching the theory of
complex numbers to students \cite{reynolds83types}. One professor
chooses rectangular coordinates ($x + iy$), while the other chooses
polar coordinates ($\alpha\cos\theta + i\alpha\sin\theta$). Happily,
after learning the basic operations on complex numbers in the two
representations, the two classes can interact because the theory of
complex numbers is invariant with respect to the choice of
representation. Reynolds formalises the idea of invariance with
respect to changes of representation by using the notion of
preservation of relations. If a binary relation $R$ relates two
representations of a concept---for example the rectangular and polar
representations of the same complex number are related---then a
program that manipulates complex numbers at a level of abstraction
above their specific representation should preserve the relation
$R$. Concretely, let $f$ be a program taking complex numbers to
complex numbers that is intended to respect invariance under change of
representation. Then if $c$ is a complex number in rectangular form,
and $c'$ is a complex number in polar form such that $R$ relates them,
which we write as $(c,c') \in R$, then it should be the case that
$(f(c), f(c')) \in R$.

We now apply Reynolds' relational approach to defining invariance
under changes of representation to our geometric setting. Following
Reynolds, we express the fact that two tuples in $\mathbb{R}^2$ are
representatives of the same point under change of origin by defining a
family of binary relations over $\mathbb{R}^2$. Our family of binary
relations is parameterised by a particular choice of origin, and by
quantifying over all choices of origin we will be able to ensure that
our programs are invariant with respect to the particular choice of
origin. We define a $\mathbb{R}^2$-parameterised family of binary
relations $\{ R_{\vec{o}} \subseteq \mathbb{R}^2 \times \mathbb{R}^2
\}_{\vec{o} \in \mathbb{R}^2}$ as follows:
\begin{displaymath}
  R_{\vec{o}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{o} \}
\end{displaymath}

@@

We have seen how Reynolds' notion of preservation of relations can be
used to define when it is the case that programs manipulating
representatives of points and vectors make geometric sense. Reynolds
further insight was to see how preservation of relations can
automatically be established by following a typing discipline. In
Reynolds' case, the typing discipline was the polymorphic
$\lambda$-calculus where new types can be constructed by universal
quantification over all types. Universal type quantification is
commonly written $\forall \alpha. A$, which denotes a type that may be
instantiated with any other type $B$ by replacing the type variable
$\alpha$ with $B$ in $A$.

@@

Equipped with quantification over the group $T_2$ of invertible affine
maps, and the parameterised types $\mathsf{vec}\langle \cdot \rangle$,
we can now express origin invariance properties as types. In the
previous section, we discussed functions $f : \mathbb{R}^2 \times
\mathbb{R}^2 \to \mathbb{R}^2$ that have the property that they
preserve all changes of origin. We can now express this property as a
type:
\begin{displaymath}
  \mathrm{f} : \forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \times \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle
\end{displaymath}
FIXME: forward references...

% \subsection{Affine Combination of Points}
% \label{sec:affine-combination}

% As we stated in the previous section, the primitive operation of
% affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
% \lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
% affine combination can be interpreted as interpolation between the
% points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
% combination of points to our calculus, with the following typing and
% intended denotation.
% \begin{displaymath}
%   \begin{array}{l}
%     \mathrm{affineCombination} :\\
%     \hspace{1.5cm}\forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \\
%     \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ t\ (x_2,y_2) = \\
%     \hspace{3cm}((1-t)x_1 + tx_2, (1-t)y_1 + ty_2)
%   \end{array}
% \end{displaymath}
% By defining the primitive function $\mathrm{affineCombination}$ to
% take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
% we are taking the affine combination of two representatives of points,
% and not just an arbitrary linear combination. Note that the intended
% denotation in the second line does not take any argument corresponding
% to the universally quantified $F$ in the type. In our calculus,
% geometric group expressions are erased at runtime.

% % In Linear Algebra notation, we could have written the right-hand side
% % as $(1-t)\vec{v_1} + t\vec{v_2}$, where $\vec{v_1} = (x_1,y_1)$ and
% % $\vec{v_2} = (x_2,y_2)$.

% \begin{example}
%   With just the single primitive operation of affine combination, we
%   can already express some useful programs. For example, the
%   evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves with
%   two endpoints and a single control point) can be expressed using the
%   affine combination primitive and three steps of De Casteljau's
%   algorithm:
%   \begin{displaymath}
%     \begin{array}{l}
%       \mathrm{quadraticB\acute{e}zier} :\\
%       \hspace{0.5cm}\forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \\
%       \mathrm{quadraticB\acute{e}zier}\ [F]\ p_0\ p_1\ p_2\ t = \\
%       \quad
%       \begin{array}{l@{\hspace{-0.002cm}}l}
%         \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ t\ p_1) \\
%         & t \\
%         & (\mathrm{affineCombination}\ [F]\ p_1\ t\ p_2)
%       \end{array}
%     \end{array}
%   \end{displaymath}
%   For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
%   application $\mathrm{quadraticB\acute{e}zier}\ p_0\ p_1\ p_2\ t$ for
%   $t \in [0,1]$ gives us the point on the curve at ``time'' $t$. Note
%   that the type of $\mathrm{quadraticB\acute{e}zier}$ tells us
%   immediately that it preserves all invertible affine maps.
%   %FIXME: forward ref to ``free theorem''.
% \end{example}

\subsection{Vector Space Operations}
\label{sec:vector-space-ops}

\textbf{FIXME}Affine combination only provides us with an operation on
points. For the purposes of computational geometry, we are also
interested in operations on the vectors representing offsets between
points. We now examine the correct types to assign to the vector space
operations of addition of vectors, negation of vectors, multiplication
by a scalar and the zero vector.

In our discussion in the previous section \textbf{?FIXME?}, we
observed that vector operations are not invariant under change of
origin. The obvious type assignment for vector addition is therefore:
\begin{displaymath}
  (+) : \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle
\end{displaymath}
(we write binary operators intended to be used infix in parentheses
when not appearing in infix position, following the Haskell
syntax.)

However, it is possible to give vector addition a more precise type
that describes its effect on the chosen origin. 
% We now introduce a new
% sort $\Transl(2)$ of translations. Translations form an abelian group,
% which we write additively, using $+$,$-$ and $0$, and we use lower
% case $t$, $t_1$, etc.~for $t$ranslation variables. Invertible affine
% maps are constructed from invertible linear maps and translations, so
% given a $B : \GL(2)$ and a $t : \Transl(2)$, we can form an invertible
% affine map $B \ltimes t : \GA(2)$. Conversely, given $F : \GA(2)$, we
% have an invertible linear map $\pi_L(F) : \GL(2)$, and a translation
% $\pi_T(F) : \Transl(2)$. Thinking of elements of $\GA(2)$ as frames,
% $\pi_L$ returns the basis of the frame, and $\pi_T$ returns the
% origin.
%
% We can now give vector addition a more general type, which
% demonstrates how adding vectors has an effect on translations:
\begin{displaymath}
  (+) : \forall t_1, t_2 \mathord: T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2 \rangle \to \mathsf{vec}\langle t_1 + t_2 \rangle
\end{displaymath}
We can also negate vectors, yielding a vector which points in the
opposite direction. Negation negates translations:
\begin{displaymath}
  \mathrm{negate} : \forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle -t \rangle
\end{displaymath}
With the primitive operations of addition and negation of vectors, we can define the derived
operation of vector subtraction with the following type:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall t_1,t_2 \mathord:T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2\rangle \to \mathsf{vec}\langle t_1 - t_2 \rangle \\
    (-)\ [B]\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
  \end{array}
\end{displaymath}

Given two points that have been constrained to be invariant with
respect to the same change of origin---i.e.~two values of type
$\mathsf{vec}\langle t \rangle$---we can compute their offset, which
is a vector expressed with respect to the null change of origin. The
offset operation is a special case of vector subtraction:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle 0 \rangle \\
    \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
Similarly, the type of the vector addition operation can be
specialised to the case of moving a point by an offset represented by
a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle t \rangle \\
    \mathrm{moveBy}\ [t]\ p\ v = p + v
  \end{array}
\end{displaymath}
The types we assign to the remaining vector space primitives, the zero
vector and multiplication by a scalar, do not describe any interesting
effect on translations%\footnote{but they could...}
. The zero vector has the following type.
\begin{displaymath}
  \mathrm{0} : \mathsf{vec}\langle 0 \rangle
\end{displaymath}
Intuitively, there cannot be a zero vector invariant under any change
of origin, because this would entail picking a privileged origin. For
multiplication of a vector by a scalar, we assign the following type,
which is again not invariant under change of origin.
\begin{displaymath}
  (*) : \mathsf{real} \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle
\end{displaymath}

\begin{example}
  The vector space operators and the properties that follow from their
  types allow us to establish a useful type isomorphism. Consider
  functions with types conforming to the following schema:
  \begin{displaymath}
    \tau_{n} \isDefinedAs \forall t\mathord:T_2.\ \mathsf{vec}\langle t \rangle^{n+1} \to \mathsf{real}
  \end{displaymath}
  We take $n$ to be a natural number, and $A^n$ for some type $A$
  denotes the $n$-ary tupling of $A$ with itself: i.e.~$A \times
  ... \times A$, with $A$ repeated $n$ times.
% The notation $A[-]$
%   denotes a type with a ``hole'' that can be filled with an
%   expression. In this case, we require that $A[e]$ be a well-formed
%   type when $e$ has sort $\GA(2)$.

  Just by looking at the types $\tau_{n}$, we know that their
  inhabitants will be invariant under change of origin, due to the
  quantification over all $t$ in $T_2$. Therefore, we may as well pick
  one of the input points as the origin and assume that all the other
  points are defined with respect to this chosen origin. This is
  similar to the common mathematical step of stating that, ``without
  loss of generality'', we may pick some point in a description of a
  problem to be the origin, as long as the problem statement is
  invariant under translation.

  The types $\tau_{n}$ are isomorphic to the corresponding types
  $\sigma_{n}$, defined below. Inhabitants of the types
  $\sigma_{n}$ take one fewer points (now seen as vector offsets
  from some origin), but are invariant only under change of basis.
  \begin{displaymath}
    \sigma_{n} \isDefinedAs \mathsf{vec}\langle 0 \rangle^n \to \mathsf{real}
  \end{displaymath}
  We demonstrate this isomorphism formally below in
  \autoref{sec:FIXME}.

@@

\subsection{Generalising from Origin Invariance}

Points for consideration in terms of language design:
\begin{itemize}
\item Knowledge of the algebraic laws by the type checker are
  essential for type-checking the examples above
\item The relational interpretation gives us 
\end{itemize}

>>>>

We motivate this paper's goal of investigating algebraically-indexed
types and their relational interpretations by presenting several
examples. Our main set of examples comprise a novel type system for
writing programs that manipulate geometric data, while maintaining
geometric invariants through the types. \fixme{and the information
  flow examples}

\subsection{2-Dimensional Geometry and Origin Invariance}

When writing programs that manipulate geometric data, the basic data
structure is the $n$-tuple of real numbers. In the 2-dimensional case,
tuples $\vec{v} = (x,y)$ may be called upon to represent both
\emph{points} -- offsets from some origin -- and \emph{vectors} --
offsets in their own right. The notions of point and vector are very
different despite their common representation. By representing a point
concretely as an offset from an origin, we inadvertently fix that
origin as privileged.

Putting this in concrete terms, if we have two origins, $(0,0)$ and
$(10,20)$, then the tuple $(1,1)$ with respect to the first origin and
$(11,21)$ with respect to the second origin represent the \emph{same}
point. It is merely an artifact of our representation of points as
pairs of numbers that these appear to be different. When we write
programs that manipulate data that represents points, we ought to
follow the general principle that our programs are invariant with
respect to changes in the choice of origin. On the other hand, vectors
are not intended to be regarded as invariant under change of choice of
origin. Vectors represent offsets, or movements, and the vector
$(0,0)$ always represents the zero offset.
% However, they are intended to be invariant under change of basis: the
% vector $(2,3)$, with respect to the basis $(1,0),(0,1)$, and the
% vector $(1,1)$, with respect to the basis $(2,0),(0,3)$, should be
% regarded as the same vector. Invariance under change of basis also
% applies to the representation of points. We therefore refine the
% principle of invariance under change of origin for points to also
% include invariant under change of basis for points and vectors.

@@

Invariance under change of representation immediately recalls
Reynolds' fable concerning two professors teaching the theory of
complex numbers \cite{reynolds83types}. One professor chooses
rectangular coordinates ($x + iy$), while the other chooses polar
coordinates ($\alpha\cos\theta + i\alpha\sin\theta$). Happily, after
learning the basic operations on complex numbers in the two
representations, the two classes can interact because the theory of
complex numbers is invariant with respect to the choice of
representation. Reynolds formalises the idea of invariance with
respect to changes of representation by using the notion of
preservation of relations. If a binary relation $R$ relates two
representations of a concept---for example the rectangular and polar
representations of the same complex number are related---then a
program that manipulates complex numbers at a level of abstraction
above their specific representation should preserve the relation
$R$. For example, if $f$ be a program taking complex numbers to
complex numbers that is intended to respect invariance under change of
representation. Then if $c$ is a complex number in rectangular form,
and $c'$ is a complex number in polar form such that $R$ relates them,
which we write as $(c,c') \in R$, then it should be the case that
$(f(c), f(c')) \in R$.

We now apply Reynolds' relational approach to defining invariance
under changes of representation to the geometric setting. Following
Reynolds, we express the fact that two tuples in $\mathbb{R}^2$ are
representatives of the same point under change of origin by defining a
family of binary relations over $\mathbb{R}^2$. Our family of binary
relations is parameterised by a particular choice of origin, and by
quantifying over all choices of origin we will be able to ensure that
our programs are invariant with respect to the particular choice of
origin. We define a $\mathbb{R}^2$-parameterised family of binary
relations $\{ R_{\vec{o}} \subseteq \mathbb{R}^2 \times \mathbb{R}^2
\}_{\vec{o} \in \mathbb{R}^2}$ as follows:
\begin{displaymath}
  R_{\vec{o}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{o} \}
\end{displaymath}

@@

Reynolds further insight was to see how preservation of relations by
the denotational interpretations of programs can automatically be
established by following a typing discipline. In Reynolds' case, the
typing discipline was the polymorphic $\lambda$-calculus where new
types can be constructed by universal quantification over all
types. Universal type quantification is commonly written $\forall
\alpha. A$, which denotes a type that may be instantiated with any
other type $B$ by replacing the type variable $\alpha$ with $B$ in
$A$.

@@

Equipped with quantification over the group $T_2$ of changes of
origin, and the parameterised types $\mathsf{vec}\langle \cdot
\rangle$, we can now express origin invariance properties as types. In
the previous section, we discussed functions $f : \mathbb{R}^2 \times
\mathbb{R}^2 \to \mathbb{R}^2$ that have the property that they
preserve all changes of origin. We can now express this property as a
type:
\begin{displaymath}
  \mathrm{f} : \forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \times \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle
\end{displaymath}
FIXME: forward references...

\subsection{Affine and Vector Operations}
\label{sec:affine-vector-ops}

As we stated in the previous section, the primitive operation of
affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
\lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
affine combination can be interpreted as interpolation between the
points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
combination of points to our calculus, with the following typing and
intended denotation:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affineCombination} :\forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{real} \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \\
    \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ r\ (x_2,y_2) = \\
    \hspace{3cm}((1-r)x_1 + tx_2, (1-r)y_1 + ty_2)
  \end{array}
\end{displaymath}
By defining the primitive function $\mathrm{affineCombination}$ to
take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
we are taking the affine combination of two representatives of points,
and not just an arbitrary linear combination. The intended denotation
in the second line does not take any argument corresponding to the
universally quantified $t$ in the type. Type-level variables are
erased at runtime.

% In Linear Algebra notation, we could have written the right-hand side
% as $(1-t)\vec{v_1} + t\vec{v_2}$, where $\vec{v_1} = (x_1,y_1)$ and
% $\vec{v_2} = (x_2,y_2)$.

\begin{example}
  The evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves
  with two endpoints and a single control point) can be expressed
  using the affine combination primitive and three steps of De
  Casteljau's algorithm:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{quadB\acute{e}zier} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{real} \to \mathsf{vec}\langle t \rangle \\
      \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\
      \quad
      \begin{array}{l@{\hspace{-0.002cm}}l}
        \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ s\ p_1) \\
        & s \\
        & (\mathrm{affineCombination}\ [F]\ p_1\ s\ p_2)
      \end{array}
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
  application $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ for $s
  \in [0,1]$ gives us the point on the curve at ``time'' $s$. Note
  that the type of $\mathrm{quadB\acute{e}zier}$ tells us immediately
  that it preserves all changes of origin
\end{example}

Affine combination only provides us with an operation on points. We
are also interested in operations on the vectors representing offsets
between points. We now examine the correct types to assign to the
vector space operations of addition of vectors, negation of vectors,
multiplication by a scalar and the zero vector. The typings of these
operations will make use of the abelian group structure of the sort of
translations $T_2$.

Vector operations are not invariant under change of origin. The
obvious type assignment for vector addition is therefore:
\begin{displaymath}
  (+) : \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle
\end{displaymath}
(we write binary operators intended to be used infix in parentheses
when not appearing in infix position, following the Haskell
syntax.)

However, it is possible to give vector addition a more precise type
that describes its effect on the chosen origin. 
\begin{displaymath}
  (+) : \forall t_1, t_2 \mathord: T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2 \rangle \to \mathsf{vec}\langle t_1 + t_2 \rangle
\end{displaymath}
We can also negate vectors, yielding a vector which points in the
opposite direction. Negation negates translation arguments:
\begin{displaymath}
  \mathrm{negate} : \forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle -t \rangle
\end{displaymath}
With the primitive operations of addition and negation of vectors, we
can define the derived operation of subtraction:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall t_1,t_2 \mathord:T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2\rangle \to \mathsf{vec}\langle t_1 - t_2 \rangle \\
    (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
  \end{array}
\end{displaymath}

Given two points that have been constrained to be invariant with
respect to the same change of origin---i.e.~two values of type
$\mathsf{vec}\langle t \rangle$---we can compute their offset, which
is a vector expressed with respect to the null change of origin. The
offset operation is a special case of vector subtraction:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle 0 \rangle \\
    \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
The type of the vector addition operation can be specialised to the
case of moving a point by a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle t \rangle \\
    \mathrm{moveBy}\ [t]\ p\ v = p + v
  \end{array}
\end{displaymath}
The types we assign to the remaining vector space primitives, the zero
vector and multiplication by a scalar, do not describe any interesting
effect on translations%\footnote{but they could...}
. The zero vector has the type:
\begin{displaymath}
  \mathrm{0} : \mathsf{vec}\langle 0 \rangle
\end{displaymath}
Intuitively, there cannot be a zero vector invariant under any change
of origin, because this would entail picking a privileged origin. For
multiplication of a vector by a scalar, we assign the following type,
which is again not invariant under change of origin.
\begin{displaymath}
  (*) : \mathsf{real} \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle
\end{displaymath}

\begin{example}
  The vector space operators and the properties that follow from their
  types allow us to establish a useful type isomorphism. Consider
  functions with types following the schema:
  \begin{displaymath}
    \tau_{n} \isDefinedAs \forall t\mathord:T_2.\ \underbrace{\mathsf{vec}\langle t \rangle \to ... \mathsf{vec}\langle t \rangle}_{n+1\textrm{ times}} \to \mathsf{real}
  \end{displaymath}
  Just by looking at the types $\tau_{n}$, we know that their
  inhabitants will be invariant under change of origin, due to the
  quantification over all $t$ in $T_2$. Therefore, we may as well pick
  one of the input points as the origin and assume that all the other
  points are defined with respect to this chosen origin. This is
  similar to the common mathematical step of stating that, ``without
  loss of generality'', we may pick some point in a description of a
  problem to be the origin, as long as the problem statement is
  invariant under translation.

  The types $\tau_{n}$ are isomorphic to the corresponding types
  $\sigma_{n}$:
  \begin{displaymath}
    \sigma_{n} \isDefinedAs \underbrace{\mathsf{vec}\langle 0 \rangle \to ... \mathsf{vec}\langle 0 \rangle}_{n\textrm{ times}} \to \mathsf{real}
  \end{displaymath}
  We demonstrate this isomorphism formally in \autoref{sec:FIXME}.

@@

\subsection{Generalising Origin Invariance}
\label{sec:motivation-generalising}

\begin{enumerate}
\item Basis invariance
\item Orthogonal transform invariance for dot product
\item Rotational invariance (the type of the cross-product)
\item Scale invariance (units of measure)
\end{enumerate}

In some cases, we can use the invariance as a programming tool
(e.g. type isomorphisms), but at others we can only use it from the
outside.

\subsection{Information Flow}
\label{sec:information-flow}

\begin{enumerate}
\item Think of information flow as passing around unforgable tokens
  \begin{enumerate}
  \item Give an example (given a piece of data protected by a token
    and a token, can we extract the data?)
  \item Need to make sure that tokens are unforgable
  \end{enumerate}
\item Could either enforce unforgability of tokens by some dynamic
  means, or by static type-based means
  \begin{enumerate}
  \item Leave the dynamic method as vague
  \item Ingredients of the static method: run-time tokens are just
    dummy values ($*$), the token type is parameterised by logical
    formulae, and the relational meaning is derived from an assignment
    of boolean values to atoms.
  \item Explain this as invariance under changes of satisfaction of
    the participants. Relate to descriptions of logics as rules that
    don't care about truth, only its preservation.
  \end{enumerate}
\item Cite Sabelfeld and Sands: ``A PER Model of Secure Information
  Flow in Sequential Programs''
\end{enumerate}
